\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-25 nobuild/jss_revision/response.tex

\usepackage{geometry}
\geometry{margin=1.1in}

\setlength{\parskip}{1ex}
\setlength{\parindent}{0ex}

\begin{document}

{\bfseries
\begin{center}
  Manuscript JSS2717\\
 ``sparseHessianFD: An R Package for Estimating Sparse
 Hessian Matrices''\\
 \vspace{\baselineskip}
 Response to editor and reviewer
\end{center}
}
\vspace{\baselineskip}

I would like to express my sincere appreciation for the time and effort that
was invested in the review.  It is obvious to me that the reviewer went through
the paper and code in great detail.  I am happy to write that I was
able to incorporate a large majority of
the suggestions into the the revised paper and package. For the
others, I hope that you will accept my
explanations and excuses.  Of course, if there are any strong
objections to any of my responses, I am happy
to consider further changes in advance of publication.

@@

(error messages related to algorithm.sty removed).

I don't remember if there is an easy way to distribute ``special'' sty files,
or if you must force the user to install them. If the sty is not really
critical it might be better to omit it.

\end{revQuote}

\begin{response}
Because \pkg{sparseHessianFD} passes CRAN check, I suspect that the
error is a result of the \pkg{algorithm} package
not being present in the reviewer's LaTeX installation.  It is a standard
package in the MacTeX distribution, but I cannot speak with authority about other
platforms. It should be easily installed through CTAN.

@@

I would prefer to continue to use the \pkg{algorithm} package, but I
could try to find another way to format the algorithms if there is a
strong objection.
\end{response}

\item \begin{revQuote}
I am not familiar with reference classes, but for most users I think it
should not be important that the package uses them. However, as a user, I
find it disconcerting that sparseHessianFD() is used as a (constructor)
function (eg. p12, and in the example in ?sparseHessianFD) but is not
documented in the help with 'usage' and 'arguments' as is usual for
functions. There seems to be little guidance in the usual R places about how
reference classes should be documented. Perhaps you could seek some guidance
from the community on this point.
\end{revQuote}

\begin{response}
The help page for \pkg{sparseHessianFD} now includes explicit
instructions for the initializer.  It is correct that there is
little guidance in how to document reference classes, and I agree that
there should be a way for the
existing mechanism to include documentation for the initializer to
an RC.  But it's hard to do it using Roxygen2 in way that also passes
CRAN checks. So I just ``hard coded'' a description of the initializer
into the details section of the sparseHessianFD-class documentation. I
could not find a more elegant way of doing it.  
\end{response}

@@

  \end{revQuote}

\begin{response}

Here I am combining my responses to items \ref{it:complex1} and
\ref{it:complex2} together.

  The complex step function is now implemented in the package, and is
  discussed briefly in Section 2.5 of the paper.  It is 
  correct that ``most R primitives work with complex numbers,''
  but there are many commonly used functions that do not, including 
  \func{log1p}, \func{gamma}, and the probability distribution
  functions. This limitation, combined with the requirement that the function be
  holomorphic, and my own inexperience with the complex step method, make
  me hesitant to promote it too heavily in the paper. But it is a nifty
  trick, and I am happy that it is now included in the package.

In terms of accuracy, the paper now includes comparisons between the true
Hessian and the two \class{sparseHessianFD} methods (finite differences
and complex step). To maintain the focus of the paper on
\pkg{sparseHessianFD}, I am not including an assessment of the accuracy of
\pkg{numDeriv}.  However, for the
timing comparisons in Table 4, I did
change the baseline \pkg{numDeriv} method from \func{hessian} with the
\method{Richardson} method, to \func{jacobian} with both the
\method{simple} and 
\method{complex} methods.

@@

\begin{response}
  Fixed.  Thanks.
\end{response}

\item \begin{revQuote}
 The file replication.R does not set the RNG seed. This may not be too
important if only times are generated, but will be if resulting values are
included.

  \end{revQuote}

\begin{response}
  I added a \func{set.seed} statement near the top of
  \filename{replication.R}, but it is correct that because only
  times are reported, this does not matter much.
\end{response}

@@

 \item Try to assess how much of the speedup is due to reduce computing demand
and how much is due to different memory demand. (I had the impression in the
larger problems with numDeriv that my computer started to use swap space,
which resulted in a big slow down.)

 \item Consider implementing a complex step method, and do a comparison.

 \item  Assess the difference when multiple CPUs are used. (run.par == FALSE vs
run.par == TRUE)
\end{itemize}
  \end{revQuote}

  \begin{response}
I appreciate the editor's comment that these suggestions might be
better left for a future paper.  But they are reasonable points that I
would like to address in turn.
    \begin{itemize}
    \item This is discussed in Section 3.1.  In general, it really is all about the sparsity pattern, and
      specifically, the how few partitions are needed to group the
      variables.  The ``simple sparseness'' comes into play because it
      is more likely the number of partitions will be small with the
      the proportion of structural zeros is large. The \emph{arrangement} of the non-zeros is
      related to the ordering of the variables, but the permutation
      handles that during the initialization of the
      \class{sparseHessianFD} argument.  For example, variables
      arranged in a way that the Hessian has a block-arrow pattern are
      permuted to get a Hessian with a banded pattern instead.

\item I would like to defer this topic for future reserach, mainly
  because I have no idea how to do this in a generalizable way.  I think
  it has a lot to do with the architecture of the computer
  itself.  For example, my computer has 64 GB of RAM, so memory usage
  is not a binding constraint, and swap space never comes into play.

@@

  Figure 4 does break down the components of the computation
  into the time it takes to reserve memory (initialization), partition
  the variables, and compute gradients themselves.  Perhaps that is
  the kind of comparison you had in mind.

\item The package now implements the complex step method, and comparisons of
  accuracy and computation time are included in the paper.  Thank you
  for this suggestion.

\item Again, I would like to defer this topic for future research. In
  theory, there should be no difference if parallelization is done
  across distributed notes.  It would be as if each run were completed
  on a separate computer.  In a shared memory architecture, there
  could be conflicts in terms of processor demand and memory access,
  but this is true for any application.
    \end{itemize}

@@

  \end{response}

\end{enumerate}

Again, thank you for the helpful comments.  The paper is greatly
improved by them.

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.tex

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of  ``structural''  zeros
(Hessian matrix elements that are always zero, regardless of the value at which
function is estimated) is high.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

>>>>

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of  ``structural''  zeros
(matrix elements that are always zero, regardless of the value at which
function is estimated) is high.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.
